---
title: "Geospatial Network Analysis of the Cook County Medical Examiner’s Homicide Data"
author: "Daniel P. Hall Riggins, MD"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

## Methods

### Analysis

I carried out analysis in R with the following dependencies:

```{r message=FALSE, warning=FALSE}
# Orchestration
library(targets)
# Generic Data Wrangling
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
# Mapping
library(ggmap)
library(sf)
library(mapdeck)
# Network Analysis
library(tidygraph)
# User-defined functions
source("R/describing_ccme.R")
source("R/visualizing_ccme.R")
source("R/wrangling_archive.R")
source("R/wrangling_esri.R")
```

```{r}
sessionInfo()
```

I orchestrated my analytic pipeline using the {[targets](https://docs.ropensci.org/targets/)} package. Here is a visualization of the full pipeline:

```{r}
tar_visnetwork()
```

I obtained the following raw data sources:

| Dataset                         | Source                       | URL                                                                                                |
|----------------|----------------|-----------------------------------------|
| CCME Case Archive               | Cook County Medical Examiner | <https://datacatalog.cookcountyil.gov/Public-Safety/Medical-Examiner-Case-Archive/cjeq-bs86>       |
| Cook County Geographic Boundary | Cook County GIS              | <https://hub-cookcountyil.opendata.arcgis.com/datasets/ea127f9e96b74677892722069c984198_1/explore> |
| US Zip Code Areas               | Esri                         | <https://www.arcgis.com/home/item.html?id=8d2012a2016e484dafaac0451f9aea24>                        |

#### Wrangling CCME Cases

This section describes how I transformed a table of CCME cases into a graph with zip codes acting as nodes and cases serving as edges. First, I read in CCME Case Archive using my function `read_ccme_archive_raw()`. It is so verbose because I had to provide a full type specification due to `readr::read_csv()` incorrectly auto-inferring the types:

```{r}
read_ccme_archive_raw
```

The raw case archive looks like:

```{r}
glimpse(tar_read(ccme_archive_raw))
```

Next, I converted the raw cases into a table of edges with my function `wrangle_ccme_homicide_edges()`. This function uses the {janitor} package's `clean_names()` function to remove spaces from variable names, filters out cases without zip codes listed or withour valid zip codes, filters for only cases in the HOMICIDE category, and converts any 9-digit zip codes into corresponding 5-digit zip codes.

```{r}
wrangle_ccme_homicide_edges
```

The edges table looks like:

```{r}
glimpse(tar_read(ccme_homicide_edges))
```
Next, the edge table can be converted into a graph object with my function `wrangle_ccme_homicide_graph()`, which also adds measures of centrality and a grouping classification using functions from the {tidygraph} package. Notice that degree and betweenness are determined using the directed version of the graph, while closeness and grouping are determined with the undirected version. The harmonic version of the `centrality_closeness` function allows for graphs that have unconnected components.

```{r}
wrangle_ccme_homicide_graph
```
```{r}
tar_read(ccme_homicide_graph)
```
For visualization purposes, we'll want to perform some joins on the node table, we can isolated using the sequence:

```{r eval=FALSE}
tar_read(ccme_homicide_graph) |> activate(nodes) |> as_tibble
```

#### Wrangling Geographic Features

In order to map our network, we need data on the geographic boundaries of each zip code, which we obtain from the Esri dataset. However, we don't want to store this entire dataset in persistent memory since it encompasses every zip code in the entire country. Instead we'll subset to just Cook County zip codes with my function `wrangle_cook_county_zip_code_boundaries()`, which first filters to zip codes in Illinois, then filters to zip code boundaries that geographically intersect with Cook County. The first filtering step is not strictly necessary, but reduces the computational cost of the second step.

```{r}
wrangle_cook_county_zip_code_boundaries
```
```{r}
esri_zip_codes <- st_read(tar_read(esri_zip_code_boundaries_file))
glimpse(esri_zip_codes)

ccme_zip_codes <- semi_join(esri_zip_codes, ccme_homicide_nodes, by = c("ZIP_CODE" = "name"))

tar_load(cook_county_homicide_vis_nodes)
```


Here is the boundary of Cook County in a simple plot:

```{r}
plot(select(tar_read(cook_county_boundary), OBJECTID_1))
```
Here are the zip codes that intersect:

```{r}
plot(tar_read(cook_county_zip_code_boundaries))
```

#### Wrangling nodes and edges for visualization:

The full homicide graph includes zip code nodes that do not intersect with Cook County. I have included these nodes and their edges when calculating graph metrics, but for simplified visualization, we will only include those intersecting with Cook County. The {mapdeck} visualization library requires graph data to be structured differently than {tidygraph}. We'll generate the corresponding edge table with my function `wrangle_cook_county_homicide_vis_edges()`, which collapses all edges with the same from-to pair (weighting to their overall sum), joins edges to the coordinates for the geographic centers of each from/to zip code, and adds the from/to nodes' graph metrics to the edge table:

```{r}
wrangle_cook_county_homicide_vis_edges
```
We'll generate the corresponding node table with an inner join between the geographic boundary and nodes tables:

```{r}
inner_join(cook_county_zip_code_boundaries, ccme_homicide_nodes, by = c("ZIP_CODE" = "name"))
```

#### Mapping 

We'll generate maps of centrality measures using my function `map_ccme_centrality()`, where argument `centrality_var` is the string name for the metric of interest. Sidenote: If wanting to reproduce these maps in your own environment, you will need to follow {mapdeck} instructions for setting up an API token with Mapbox.

```{r}
map_ccme_centrality
```

We'll generate a map of algorithm-identified node groups with my function `map_ccme_neighborhoods()`, which pre-filters for the top 5 largest groups for simplified interpretation:

```{r}
map_ccme_neighborhoods
```


## Results

### Homicide Summary Statistics

We produce a table of summary statistics for CCME homicide cases using the sequence:

```{r message=FALSE, warning=FALSE}
tar_read(ccme_homicide_edges) |> 
        select(age, gender, race, latino) |> 
        gtsummary::tbl_summary()
```

The majority of CCME homicide cases involve 20-to-30-something, Black males.

### Homicide Network Maps:

```{r}
ccme_homicide_nodes |> 
    filter(
            in_degree_perc_rank > 0.75 |
            out_degree_perc_rank > 0.75
    ) |> 
    filter(
        in_out_diff > 0.75 |
            in_out_diff < -0.75
    )
    
```


#### Degree

Zip codes with high out-degree correspond to those with the most incidents of homicide connected outwardly to zip codes that are homes of the victims. 

```{r}
tar_read(cook_county_homicide_degree_map)
```

High out-degree zip codes concentrate around the West- and South-Sides of Chicago.

## Closeness

```{r}
tar_read(cook_county_homicide_closeness_map)
```

## Betweenness

```{r}
tar_read(cook_county_homicide_betweenness_map)
```

## Neighborhoods

```{r}
tar_read(cook_county_homicide_neighborhood_map)
```

## Conclusions
